A Switching Planner for Combined Task and Observation Planning

### 2011-05-01 subarchitectures/planner.sa/trunk/src/latex/AAAI11/dtpddl.tex
bunch of changes to smaller details.

We give an overview of the declarative first-order language DTPDDL, an
extension of PPDDL that can express probabilistic models of the
sensing consequences of acting, to quantitatively capture
unreliability in perception. There are straightforward compilations
from problems expressed in DTPDDL to flat (and factored)
representations of the underlying decision process. Although similar
to Bryce's POND input language, DTPDDL distinguishes itself by
explicitly treating state and perceptual symbols separately, and by
providing distinct declarations for operators (i.e, state model) and
senses (i.e., observation model). In this last respect, DTPDDL admits
more compact domain descriptions where sensing effects are common
across multiple operators. In detail, DTPDDL has perceptual analogues of
fluent and predicate symbols. For example, a simple {\em object search}
domain would have:

@@

We now review the DTPDDL syntax for describing an initial state
distribution, taken verbatim from PPDDL, in order to aid us in
discussing our planner. That distribution is expressed in a tree-like
structure of terms. Each term is either: (1) atomic, e.g., a state
proposition such as $(=(\pp{is-in}~\pp{box})~\pp{office})$, (2)
probabilistic, e.g., $(\pp{probabilistic}~\prob_1 (T_1) .. \prob_n
(T_n))$ where $T_i$ are conjunctive, or (3) a conjunct over
probabilistic and atomic terms. The root term is always conjunctive,
and the leaves are atomic. For example, a simplified object search
could have:\footnote{In PDDL, $(\pp{:init}~T_1..T_n)$ expresses the
conjunctive root of the tree -- i.e., the root node
$(\pp{and}~T_1..T_n)$. Also, we shall write $\prop$, rather than
$(\pp{and}~\prop)$, for conjunctive terms that contain a single atomic
subterm.}

\vspace{-1ex}
\small
\begin{tabtt}
(\=:init (= (is-in R2D2) kitchen) \+ \\
       (probabilistic \=.8 (= (is-in box) office)  \\
		      \>.2 (= (is-in box) kitchen)) \\
       (probabilistic .3 (= (is-in cup) office)  \\
		      \>.7 (= (is-in cup) kitchen))) \\
\end{tabtt}
\normalsize

@@

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
\hline
\end{tabular}
\normalsize

>>>>

We give an overview of the declarative first-order language DTPDDL, an
extension of PPDDL that can express probabilistic models of the
sensing consequences of acting, to quantitatively capture
unreliability in perception. There are straightforward compilations
from problems expressed in DTPDDL to flat state-based (and
propositionally factored) representations of the underlying decision
process. Although similar to the POND input
language~\cite{Bryce:2008}, DTPDDL distinguishes itself by explicitly
treating state and perceptual symbols separately, and by providing
distinct declarations for operators (i.e, state model) and senses
(i.e., observation model). In this last respect, DTPDDL admits more
compact domain descriptions where sensing effects are common across
multiple operators. In detail, DTPDDL has perceptual analogues of
fluent and predicate symbols. For example, a simple {\em object
search} domain would have:

@@

We now review the DTPDDL syntax for describing an initial state
distribution, taken verbatim from PPDDL, in order to aid us in
discussing our planner. That distribution is expressed in a tree-like
structure of terms. Each term is either: (1) atomic, e.g., a state
proposition such as $(=(\pp{is-in}~\pp{box})\pp{office})$, (2)
probabilistic, e.g., $(\pp{probabilistic}~\prob_1 (T_1) .. \prob_n
(T_n))$ where $T_i$ are conjunctive, or (3) a conjunct over
probabilistic and atomic terms. The root term is always conjunctive,
and the leaves are atomic. For example, a simplified object search
could have:\footnote{In PDDL, $(\pp{:init}~T_1..T_n)$ expresses the
conjunctive root of the tree -- i.e., the root node
$(\pp{and}~T_1..T_n)$. Also, we shall write $\prop$, rather than
$(\pp{and}~\prop)$, for conjunctive terms that contain a single atomic
subterm.}

\vspace{-1ex}
\small
\begin{tabtt}
(\=:init (= (is-in Robot) kitchen) \+ \\
       (probabilistic \=.8 (= (is-in box) office)  \\
		      \>.2 (= (is-in box) kitchen)) \\
       (probabilistic .3 (= (is-in cup) office)  \\
		      \>.7 (= (is-in cup) kitchen))) \\
\end{tabtt}
\normalsize

@@

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in Robot)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
\hline
\end{tabular}
\normalsize

### 2011-05-01 subarchitectures/planner.sa/trunk/src/latex/AAAI11/switching.tex
bunch of changes to smaller details.

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
%% .24 & kitchen & office & office \\
%% .06 & kitchen & kitchen & office \\
%% .56 & kitchen & office & kitchen \\
%% .14 & kitchen & kitchen & kitchen \\
.7 & kitchen & \# &  kitchen\\
.3 & kitchen & \# & office \\
.8 & kitchen & office & \# \\
.2 & kitchen & kitchen & \# \\
1.0 & kitchen & \# & \# \\
\hline
\end{tabular}
\normalsize

@@

\small
\[
\begin{array}{l}
\state_0 \equiv \{(=(\pp{is-in}~\pp{R2D2})~\pp{kitchen}),\\
\;\;(=(\pp{is-in}~\pp{box})~\#), (=(\pp{is-in}~\pp{cup})~\#)\}.
\end{array}
\]
\normalsize

@@

Encapsulating state-assumptions, we annotate the problem posed during
a sequential session with an \emph{assumptive action} $\assumptiveS{i}$ for
each element $\prob_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. Here, $\assumptiveS{i}$ can be executed if no
$\assumptiveS{j}$, $j \neq i$, has been executed from the same
probabilistic term, and, either
$(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root conjunct, or
it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
We also add constraints that forbid scheduling of
assumptions about facts after actions with preconditions or effects
that mention those facts. For example, the robot cannot assume it is
plugged into a power source immediately after it unplugs itself.
%%
Executing $\assumptiveS{i}$ in a state $\state$ effects a transition
to a successor state $\state^{T_i}$ with probability $\prob_i$, and
$\state^\bot$ with probability $1 - \prob_i$. Here, $\state^{T_i}$ is
the union of $\state$ with atomic terms from $T_i$. State
$\state^\bot$ is an added sink.

We now describe the optimisation criteria used during sequential
sessions. Where $\prob_i$ is the probability that the $i^{th}$
sequenced action, $\action_i$, from a trace of state-action pairs
$\langle \state_0, \action_0,\state_1, \action_1,.., \state_N \rangle$
does not transition to $\state^\bot$, we have that the optimal trace
has value:

@@

\small
\[
\begin{array}{l}
\actions^{\circ}(.8;(=(\pp{is-in}~\pp{box})\pp{office}));\\
\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}));\\
(\pp{look}~\pp{box}~\pp{office});
(\pp{look}~\pp{cup}~\pp{kitchen});\\
(\pp{report}~\pp{box}~\pp{office}); 
(\pp{report}~\pp{cup}~\pp{kitchen})
\end{array}
\]
\normalsize

@@

\Omit{
%%
Finally, 
If an
assumption was rejected, we prohibit that sequential session from
making it again.
}

%%  the size of the state space
%% is determined by two factors: the number of non-zero states in the
%% initial belief-state, and the number of actions. Determining
%% action reachability is intractable, as is determining their relevance
%% to goal achievement.

%% As we are aiming to keep the decision theoretic session fast, we try
%% to keep the state space of the abstract problem as small as
%% possible. In a deterministic-action POMDP, the state space size is
%% determined by two factors: the number of non-zero states in the
%% initial belief-state and the number of executable actions. As
%% determining the relevance of an action for a given goal in in general
%% a hard problem, we concentrate on keeping the number of uncertain
%% facts as low as possible without losing important information about
%% the sensing goal.

In order to guarantee fast DT sessions, planning occurs for a small
abstract problem. In our setting, where the underlying decision
process has a deterministic action model, a significant cause of
problem largeness is due to the number of states that occur with
non-zero probability in the underlying belief-state. In posing a small
abstract process for a DT session, we construct an abstract
belief-state that is chiefly characterised by facts related to the
current trace. In that abstraction, instances of operator or sense
declarations whose specification mentions an excluded fact are not
available.
%%
In detail, we first construct an $(\pp{:init})$ declaration which
contains the root term, and relevant assumptions from the trace --
e.g., (Fig.\ref{fig:abstraction-b}) gives an example fragment, where
diamonds are probabilistic terms, and circles are atomic and/or
conjunctive.
%%
For each excluded atom, we compute the {\em entropy} of the {\em
relevant}
assumptions, {\em conditional} on that atom. Intuitively, lower
entropy indicates an atom gives better information about
assumptions. Facts are iteratively added to the belief-state in
increasing order according to that measure until the belief space size
reaches a predefined limit (Fig.\ref{fig:abstraction-c}).

%%  or no more atoms that reduce the entropy
%% can be added

\begin{figure}[h!]
  \centering
  \tikzstyle{tree} = [sibling distance=4.5mm]
  \tikzstyle{toplevel} = [grow'=right, sibling distance=22mm]
  \tikzstyle{seclevel} = [sibling distance=9mm]
  \tikzstyle{pnode} = [diamond, draw=black, minimum size=2.5mm]
  \tikzstyle{cnode} = [circle, draw=black, minimum size=3mm]
  \tikzstyle{assumption} = [solid, very thick, draw=black]
  \tikzstyle{selected} = [solid, draw=black]
  \tikzstyle{unused} = [densely dashed, draw=black!40]
  %% \subfloat[The initial belief-state with the assumptions made by the continual
  %%   planner in bold.]{
  %%     \label{fig:abstraction-a}
  %%     \begin{tikzpicture}[
  %%   level 3/.style={tree}]
  %%   \node[pnode, assumption] (cat) at (1,1) {} [toplevel]
  %%     child[seclevel] {node[cnode, assumption] (office) {}
  %%       child {node [pnode, assumption] (box) {}
  %%         child {node [cnode] (boxp1) {}}
  %%         child {node [cnode, assumption] (boxp2) {}}
  %%       }
  %%       child {node [pnode] (cup) {} 
  %%         child {node [cnode] (cupp1) {}}
  %%         child {node [cnode] (cupp2) {}}
  %%       }
  %%     }
  %%     child {node[cnode] (kitchen) {}
  %%       child {node [pnode] (box2) {} 
  %%         child {node [cnode] (box2p1) {}}
  %%         child {node [cnode] (box2p2) {}}
  %%       }
  %%     };
  %%  \tiny
  %%  \draw[assumption] (cat) -- (office) -- (box) -- (boxp2);
  %%  \node[above=0 of cat] {$\pp{(category room1)}$};
  %%  \node[below=0 of kitchen] {$\pp{kitchen}$};
  %%  \node[below=0 of office] {$\pp{office}$};
  %%  \node[below=0 of box] {$\pp{(is-in box)}$};
  %%  \node[below=0 of box2] {$\pp{(is-in box)}$};
  %%  \node[below=0 of cup] {$\pp{(is-in cup)}$};
  %%  \node[right=0 of boxp1] {$\pp{place1}$};
  %%  \node[right=0 of boxp2] {$\pp{place2}$};
  %%  \node[right=0 of box2p1] {$\pp{place1}$};
  %%  \node[right=0 of box2p2] {$\pp{place2}$};
  %%  \node[right=0 of cupp1] {$\pp{place1}$};
  %%  \node[right=0 of cupp2] {$\pp{place2}$};
  %%   % \node[node] (kitchen) right of (cat) {};
  %% \end{tikzpicture}}
\qquad
  \subfloat[Removing facts that are not part of an assumption.]{
    \label{fig:abstraction-b}
    \begin{tikzpicture}[
    level 3/.style={tree}]
    \node[pnode, assumption] (cat) at (1,1) {} [toplevel, unused]
      child[seclevel] {node[cnode, assumption] (office) {}
        child {node [pnode, assumption] (box) {}
          child {node [cnode, unused] (boxp1) {}}
          child {node [cnode, assumption] (boxp2) {}}
        }
        child {node [pnode, unused] (cup) {} 
          child {node [cnode, unused] (cupp1) {}}
          child {node [cnode, unused] (cupp2) {}}
        }
      }
      child {node[cnode, selected] (kitchen) {}
        child {node [pnode, selected] (box2) {} 
          child {node [cnode, unused] (box2p1) {}}
          child {node [cnode, selected] (box2p2) {}}
        }
      };
   \draw[assumption] (cat) -- (office) -- (box) -- (boxp2);
   \draw[selected] (cat) -- (kitchen) -- (box2) -- (box2p2);
   \tiny
   \node[above=0 of cat] {$\pp{(category room1)}$};
   \node[below=0 of office] {$\pp{office}$};
   \node[below=0 of box] {$\pp{(is-in box)}$};
   \node[below=0 of box2] {$\pp{(is-in box)}$};
   \node[right=0 of boxp2] {$\pp{place2}$};
   \node[right=0 of box2p2] {$\pp{place2}$};
    % \node[node] (kitchen) right of (cat) {};
  \end{tikzpicture}}
\vspace{2mm}
  \subfloat[Refinement, by adding relevant facts.]{
    \label{fig:abstraction-c}
    \begin{tikzpicture}[
    level 3/.style={tree}]
    \node[pnode, assumption] (cat) at (1,1) {} [toplevel, unused]
      child[seclevel] {node[cnode, assumption] (office) {}
        child {node [pnode, assumption] (box) {}
          child {node [cnode, selected] (boxp1) {}}
          child {node [cnode, assumption] (boxp2) {}}
        }
        child {node [pnode, unused] (cup) {} 
          child {node [cnode, unused] (cupp1) {}}
          child {node [cnode, unused] (cupp2) {}}
        }
      }
      child {node[cnode, selected] (kitchen) {}
        child {node [pnode, selected] (box2) {} 
          child {node [cnode, selected] (box2p1) {}}
          child {node [cnode, selected] (box2p2) {}}
        }
      };
   \draw[assumption] (cat) -- (office) -- (box) -- (boxp2);
   \draw[selected] (box) -- (boxp1);
   \draw[selected] (cat) -- (kitchen) -- (box2) -- (box2p2);
   \draw[selected] (box2) -- (box2p1);
   \tiny
   \node[above=0 of cat] {$\pp{(category room1)}$};
   \node[below=0 of office] {$\pp{office}$};
   \node[below=0 of box] {$\pp{(is-in box)}$};
   \node[below=0 of box2] {$\pp{(is-in box)}$};
   \node[right=0 of boxp1] {$\pp{place1}$};
   \node[right=0 of boxp2] {$\pp{place2}$};
   \node[right=0 of box2p1] {$\pp{place1}$};
   \node[right=0 of box2p2] {$\pp{place2}$};
    % \node[node] (kitchen) right of (cat) {};
  \end{tikzpicture}}

  \caption{Abstract belief-state, and refinement.}
\label{fig:abstraction}
\end{figure}

% When limiting the number of actions, we aim to include those that are
% possibly required to a) reach the goal or b) to directly observe an
% atom that is part of the abstract belief state. In general, checking
% this for an action is \textsc{PSPACE} complete for \textsc{STRIPS}
% planning.

% We approximate the set of relevant actions as follows: Let
% \relevantAtoms\ be the set of atoms that either occur in the goal
% condition \poss(\switchAction) or in a precondition \poss($a$) of an
% action $a$ that can directly sense any atom in the abstract belief
% state. We construct a \emph{relaxed planning graph} with
% \relevantAtoms\ as goal atoms from which we extract a relaxed
% plan.

%% In the first step, we remove all facts that are not part of an
%% assumption (Fig. \ref{fig:abstraction-b}). At this point, the session
%% would proceed in an abstraction of the environment that does not
%% contain $\pp{place1}$, the $\pp{cup}$ or a $\pp{kitchen}$. 
%% %%
%% In a second step, we iteratively refine the relaxed declaration by
%% adding terms from the original statement of $(\pp{:init})$ while the
%% number of abstract states in $\bstate_0$ that occur with non-zero
%% probability according to that refined declaration remains of a
%% practicable size. In detail,

>>>>

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in Robot)  & (is-in box)  & (is-in cup) \\
\hline
.7 & kitchen & \# &  kitchen\\
.3 & kitchen & \# & office \\
.8 & kitchen & office & \# \\
.2 & kitchen & kitchen & \# \\
1.0 & kitchen & \# & \# \\
\hline
\end{tabular}
\normalsize

@@

\small
\[
\begin{array}{l}
\state_0 \equiv \{(=(\pp{is-in}~\pp{Robot})\pp{kitchen}),\\
\;\;(=(\pp{is-in}~\pp{box})\#), (=(\pp{is-in}~\pp{cup})\#)\}.
\end{array}
\]
\normalsize

@@

To represent state assumptions we augment the problem posed during
a sequential session with an \emph{assumptive action} $\assumptiveS{i}$ for
each element $\prob_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. Here, $\assumptiveS{i}$ can be executed if no
$\assumptiveS{j}$, $j \neq i$, has been executed from the same
probabilistic term, and, either
$(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root conjunct, or
it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
We also add constraints that forbid scheduling of
assumptions about facts after actions with preconditions or effects
that mention those facts. For example, the robot cannot assume it is
plugged into a power source immediately after it unplugs itself.
%%
Executing $\assumptiveS{i}$ in a state $\state$ effects a transition
to a successor state $\state^{T_i}$ with probability $\prob_i$, and
$\state^\bot$ with probability $1 - \prob_i$. Here, $\state^{T_i}$ is
the union of $\state$ with atomic terms from $T_i$. State
$\state^\bot$ is an added sink.

We now describe the optimisation criteria used during sequential
sessions. Let $\prob_i$ be the probability that the $i^{th}$ sequenced
action, $\action_i$, from a trace of state-action pairs
$\langle \state_0, \action_0,\state_1, \action_1,.., \state_N \rangle$
does not transition to $\state^\bot$, then the optimal
trace has value:

@@

\small
\[
\begin{array}{l}
\actions^{\circ}(.8;(=(\pp{is-in}~\pp{box})\pp{kitchen}));\\
\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{office}));\\
(\pp{look}~\pp{box}~\pp{kitchen});
(\pp{look}~\pp{cup}~\pp{office});\\
(\pp{report}~\pp{box}~\pp{kitchen}); 
(\pp{report}~\pp{cup}~\pp{office})
\end{array}
\]
\normalsize

@@

%%%%

In order to guarantee fast DT sessions, those plan in an abstract
process determined by the current trace and underlying belief-state.
%%
The abstract process posed to the DT planner is constructed by first
constraining as statically false all propositions except those which
are true with probability 1, or which are the subject of {\em
relevant} assumptions. For example, taking the above trace and
switching action {\texttt (look~box~office)}, the underlying
belief in Fig.~\ref{fig:beliefs}B would determine a fully
constrained belief given by Fig.~\ref{fig:beliefs}C.
%%
Next, static constraints are removed, one proposition at a time, until
the number of states that can be true with non-zero probability in the
initial belief of the abstract process reaches a given threshold.
%%
In detail, for each statically-false proposition we compute the {\em
entropy} of the state assumptions of the current trace {\em
conditional} on that proposition. 
%%
Let $X$ be a set of propositions and $2^X$ the powerset of $X$, then taking

\small
\[
\chi = \{\bigwedge_{x \in X'
  \cap X}x \; \land \bigwedge_{x \in X \setminus X'}\neg x \;|\; X' \in 2^X\},
\]
\normalsize

\noindent we have that $\chi$ is a set of conjunctions each of which
corresponds to one truth assignment to elements in $X$. Where
$p(\phi)$ gives the probability that a conjunction $\phi$ holds in the
belief-state of the DTPDDL process, the entropy of $X$
\emph{conditional} on a proposition $y$, written $\entropy(X|y)$, is
given by Eq.~\ref{eq:condent}.

\vspace{-1ex}
\small
\begin{equation}\label{eq:condent}
  \entropy(X|y) = \sum_{x \in \chi, y' \in \{y, \neg y\}} p(x \land y') \log_2
  \frac{p(y')}{p(x \land y')}
\end{equation}
\normalsize

A low $\entropy(X|y)$ value suggests that knowing the truth value of
$y$ is useful for determining whether or not some assumptions $X$
hold. When removing a static constraint on propositions during the
abstract process construction, $y_i$ is considered before $y_j$ if
$\entropy(X|y_i)~<~\entropy(X|y_j)$. For example, if the serial plan
assumes the box is in the kitchen, then propositions about the
contents of kitchens containing a box, e.g. {\texttt
  (=(is-in~milk)kitchen)}, are added to characterise the abstract
process' states. Taking a relevant assumption $X$ to be {\texttt
  (=(is-in~box)kitchen)}, in relaxing static constraints the following
entropies are calculated:

\small
\begin{tabtt}
.47 \= = \entropy(\=X|(=(is-in~milk)office))\+ \\
 = \entropy(X|(=(is-in~milk)kitchen)) \-\\
.97 = \entropy(X|(=(is-in~cup)office))\+\\
 = \entropy(X|(=(is-in~cup)kitchen))
\end{tabtt}
\normalsize

\noindent Therefore, the first static constraint to be relaxed is for
\texttt{(=(is-in~milk)office)}, or
equivalently \texttt{(=(is-in~milk)kitchen)}, giving a refined
abstract belief state depicted in
Fig.~\ref{fig:beliefs}A. Summarising, if for Fig.\ref{fig:beliefs}B
the DT session is restricted to belief-states with fewer than $8$
elements, then the starting belief-state of the DT session does not
mention a ``cup''.

\begin{figure}
\tiny
\begin{tabularx}{\columnwidth}{XX}
\hline
\vspace{1ex}
\scriptsize (A) Partially constrained belief & %abstract belief-state & 
\vspace{1ex}\scriptsize \hspace{-4ex}(B) Underlying DTPDDL belief \\
\vspace{-2ex} 
\begin{tabtt}\hspace{-4ex}
(\=:init (=(is-in~Robot)kitchen) \+ \\
       (\=.6(a\=nd(=(is-in~box)kitchen) \+\+ \\
        (\=.9(=(is-in~milk)~kitchen))\+\\
         .1(=(is-in~milk)office))\-\-  \\
       .4(and(=(is-in~box)office) \+\\
        (.1(=(is-in~milk)kitchen))\+\\
        .9(=(is-in~milk)office)))
\end{tabtt} &
\vspace{-2ex} 
\begin{tabtt}
\hspace{-4ex}(\=:init (=(is-in~Robot)office) \+ \\
       (\=.6(a\=nd(=(is-in~box)kitchen) \+\+ \\
        (\=.9(=(is-in~milk)kitchen))\+\\
         .1(=(is-in~milk)office))\-\-  \\
       .4(and(=(is-in~box)office) \+\\
        (.1(=(is-in~milk)kitchen))\+\\
        .9(=(is-in~milk)office))) \-\-\-\\
       (\=.6(=(is-in~cup)office) \+ \\
      .4(=(is-in~cup)kitchen))) \\
\end{tabtt} \\
\vspace{-8ex}\scriptsize(C) Fully constrained belief & \\
\vspace{-7.5ex} \begin{tabtt}
(\=:init (=(is-in~Robot)kitchen) \+ \\
       (\=.6(=(is-in~box)kitchen))) \\
\end{tabtt} & \\
\hline
\end{tabularx}
\caption{Simplified examples of abstract belief-states from
       DT sessions. \label{fig:beliefs}}
\normalsize
\end{figure}
